---
title: "Violations & Complaints"
author: "Francisco Brady"
date: "2023-04-14"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    theme: united
    toc_float:
      collapsed: false
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# setup libs
library(dplyr)
library(lubridate)
library(ggplot2)
library(tidyr)
library(janitor)
library(kableExtra)
library(sf)

## read in data 
violations <- vroom::vroom('building_violations.csv') %>% 
  mutate(`VIOLATION LAST MODIFIED DATE` = mdy(`VIOLATION LAST MODIFIED DATE`),
         `VIOLATION DATE` = mdy(`VIOLATION DATE`),
         `VIOLATION STATUS DATE` = mdy(`VIOLATION STATUS DATE`) ) 

ordinance_violations <- vroom::vroom('ordinance_violations.csv') %>%
  mutate(`HEARING DATE` = mdy(`HEARING DATE`),
         `VIOLATION DATE` = mdy_hms(`VIOLATION DATE`),
         `LAST MODIFIED DATE` = mdy_hms(`LAST MODIFIED DATE`)
         ) %>%
  # drop strangely old dates for now 
  filter(year(`VIOLATION DATE`) >= 2006, 
         year(`VIOLATION DATE`) <= year(today())) 
  # convert to regular dates 
ordinance_violations <- ordinance_violations %>%
  mutate(across(contains('DATE'), lubridate::as_date))

calls_311 <- vroom::vroom('calls_311.csv.gz', guess_max = 300) %>%
  # convert dates 
  mutate(CREATED_DATE = mdy_hms(CREATED_DATE),
         LAST_MODIFIED_DATE = mdy_hms(LAST_MODIFIED_DATE),
         CLOSED_DATE = mdy_hms(CLOSED_DATE)) %>%
  mutate(across(contains('DATE'), lubridate::as_date))
  

# ordinance_violations %>% select(contains('DATE')) %>% View
## game plan : 
# 1. read in data 
# 2. formats
# 3. time coverage 
# 4. types of complaints 
# 5. valid addresses

## helper functions 
missing_table <- 
  function(df) {
    df %>%
      summarise(across(everything(), ~ sum(is.na(.x)) / length(.x), .names = "{.col}_missing")) %>%
      pivot_longer(cols = everything()) %>%
      separate(name, into = c('variable','stat'), sep="_(?=[^_]+$)") %>%
      pivot_wider(names_from = 'stat', values_from = 'value') %>%
      mutate(pct_missing = round(100*missing, 2)) %>%
      select(variable,
             pct_missing) -> tbl
    return(tbl)
  }

# map data 
chi_sf <- tigris::places(state = 'IL') %>% filter(NAME == 'Chicago')
```

## Building Violations 

Building violations dataset from the City of Chicago. Coverage is fairly well distributed throughout the city. 

### Data Format  

```{r violations-summary}
varnames <- setNames(stack(sapply(violations, class))[2:1], c('variable', 'class'))

records <- format(nrow(violations), big.mark = ',')
# add pct missing 
summary <- violations %>% 
  missing_table() %>%
  right_join(varnames, c('variable' = 'variable')) %>%  
  # rearrange
  select('Variable' = variable,
         'Type' = class,
         'Percent Missing' = pct_missing)
  # print
summary %>% knitr::kable() %>% 
  kable_styling("striped", full_width = F) %>% 
  add_footnote(glue::glue('N = {records}')) %>%
  scroll_box(height = "300px")
```

### Date Coverage

```{r violations-time}
time <- violations %>%
  select(contains('DATE')) %>% #names
  summarise(across(.cols = everything(), 
                   # use in anonymous function to include na.rm arg
                   list(min = function(x) min(x, na.rm=TRUE), 
                        max = function(x) max(x, na.rm=TRUE)))) %>%
  pivot_longer(cols = everything()) %>%
  separate(name, into = c('variable','stat'), sep="_(?=[^_]+$)") %>%
      pivot_wider(names_from = 'stat', values_from = 'value')

# print
time %>% knitr::kable(col.names = c('Variable', 'Minimum', 'Maximum')) %>% 
  kable_styling("striped", full_width = F) 

```

### Top 50 Violations 
```{r, violations-type}
ords <-  violations %>%
  filter(!is.na(`VIOLATION ORDINANCE`)) %>%
  group_by(`VIOLATION ORDINANCE`) %>% summarise(count = n()) %>% 
  arrange(desc(count)) %>% 
  filter(row_number() <= 50) 

ords %>% knitr::kable(col.names = c('Ordinance', 'Count'), 
                      format.args = list(big.mark = ',')) %>% 
  kable_styling("striped", full_width = F) %>% 
  scroll_box(height = "300px")

```

### Geographic Coverage 

Only showing the latest year of data. 

```{r, v-geo, warning=FALSE, message=FALSE}

# st_crs(chi_sf)
v_sf <- violations %>% 
  filter(!is.na(LONGITUDE), !is.na(LATITUDE)) %>%
  mutate(year = year(`VIOLATION DATE`)) %>%
  select(year, LONGITUDE, LATITUDE) %>%
  filter(year == max(year)) %>%
  st_as_sf(coords = c("LONGITUDE","LATITUDE")) %>%
  st_set_crs(st_crs(chi_sf))

ggplot() + 
  geom_sf(data = chi_sf) +
  geom_sf(data = v_sf,
    aes(geometry = geometry), 
    color = 'cadetblue',
    alpha = .5,
    stat = "sf_coordinates"
  ) +
  theme_light() + 
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.y=element_blank(),
        axis.title.x=element_blank()) +
  labs(title = 'Building Violations', 
       caption = glue::glue('Year: {max(year(max(violations$`VIOLATION DATE`)))}'))


```

## Ordinance Violations

### Data Format  

```{r ords-summary}
varnames <- setNames(stack(sapply(ordinance_violations, class))[2:1], c('variable', 'class'))

records <- format(nrow(ordinance_violations), big.mark = ',')
# add pct missing 
summary <- ordinance_violations %>% 
  missing_table() %>%
  right_join(varnames, c('variable' = 'variable')) %>%  
  # rearrange
  select('Variable' = variable,
         'Type' = class,
         'Percent Missing' = pct_missing)
  # print
summary %>% knitr::kable() %>% 
  kable_styling("striped", full_width = F) %>% 
  add_footnote(glue::glue('N = {records}')) %>%
  scroll_box(height = "300px")
```

### Date Coverage

```{r ords-time}
time <- ordinance_violations %>%
  select(contains('DATE')) %>% #View
  summarise(across(.cols = everything(), 
                   # use in anonymous function to include na.rm arg
                   list(min = function(x) min(x, na.rm=TRUE), 
                        max = function(x) max(x, na.rm=TRUE)))) %>%
  pivot_longer(cols = everything()) %>%
  separate(name, into = c('variable','stat'), sep="_(?=[^_]+$)") %>%
  mutate(value = as_date(value)) %>%
      pivot_wider(names_from = 'stat', values_from = 'value')

# print
time %>% knitr::kable(col.names = c('Variable', 'Minimum', 'Maximum')) %>% 
  kable_styling("striped", full_width = F) 

```

### Top 50 Ordinance Violations 
```{r, ords-type}
ords <-  ordinance_violations %>%
  filter(!is.na(`VIOLATION DESCRIPTION`)) %>%
  group_by(`VIOLATION DESCRIPTION`) %>% summarise(count = n()) %>% 
  arrange(desc(count)) %>% 
  filter(row_number() <= 50) 

ords %>% knitr::kable(col.names = c('Description', 'Count'), 
                      format.args = list(big.mark = ',')) %>% 
  kable_styling("striped", full_width = F) %>% 
  scroll_box(height = "300px")

```

### Geographic Coverage 

Only showing the latest year of data. 

```{r, o-geo, warning=FALSE, message=FALSE}

# st_crs(chi_sf)
o_sf <- ordinance_violations %>% 
  filter(!is.na(LONGITUDE), !is.na(LATITUDE)) %>%
  mutate(year = year(`VIOLATION DATE`)) %>%
  select(year, LONGITUDE, LATITUDE) %>%
  filter(year == max(year, na.rm = T)) %>%
  st_as_sf(coords = c("LONGITUDE","LATITUDE")) %>%
  st_set_crs(st_crs(chi_sf))

ggplot() + 
  geom_sf(data = chi_sf) +
  geom_sf(data = o_sf,
    aes(geometry = geometry), 
    color = 'cadetblue',
    alpha = .5,
    stat = "sf_coordinates"
  ) +
  theme_light() + 
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.y=element_blank(),
        axis.title.x=element_blank()) +
  labs(title = 'Ordinance Violations', caption = glue::glue('Year: {max(year(max(ordinance_violations$`VIOLATION DATE`)))}')) 


```

## 311 Calls 

### Data Format  

```{r summary-311}
varnames <- setNames(stack(sapply(calls_311, class))[2:1], c('variable', 'class'))

records <- format(nrow(calls_311), big.mark = ',')
# add pct missing 
summary <- calls_311 %>% 
  missing_table() %>%
  right_join(varnames, c('variable' = 'variable')) %>%  
  # rearrange
  select('Variable' = variable,
         'Type' = class,
         'Percent Missing' = pct_missing)
# print
summary %>% knitr::kable() %>% 
  kable_styling("striped", full_width = F) %>% 
  add_footnote(glue::glue('N = {records}')) %>%
  scroll_box(height = "300px")
```

### Date Coverage

```{r time-311}
time <- calls_311 %>%
  select(contains('DATE')) %>% #View
  summarise(across(.cols = everything(), 
                   # use in anonymous function to include na.rm arg
                   list(min = function(x) min(x, na.rm=TRUE), 
                        max = function(x) max(x, na.rm=TRUE)))) %>%
  pivot_longer(cols = everything()) %>%
  separate(name, into = c('variable','stat'), sep="_(?=[^_]+$)") %>%
  mutate(value = as_date(value)) %>%
      pivot_wider(names_from = 'stat', values_from = 'value')

# print
time %>% knitr::kable(col.names = c('Variable', 'Minimum', 'Maximum')) %>% 
  kable_styling("striped", full_width = F) 

```

### Top 50 Call Types

Excluding information only 311 calls.

```{r, type-311}
calls <-  calls_311 %>%
  filter(!grepl('INFORMATION ONLY', SR_TYPE)) %>%
  group_by(SR_TYPE) %>% 
  summarise(count = n()) %>% 
  arrange(desc(count)) %>% 
  filter(row_number() <= 50) 

calls %>% knitr::kable(col.names = c('Call Type', 'Count'), 
                      format.args = list(big.mark = ',')) %>% 
  kable_styling("striped", full_width = F) %>% 
  scroll_box(height = "300px")

```

### Geographic Coverage: All Calls

Only showing the latest year of data. 

```{r, geo-311, warning=FALSE, message=FALSE}

# st_crs(chi_sf)
c_sf <- calls_311 %>% 
  filter(!is.na(LONGITUDE), !is.na(LATITUDE)) %>%
  mutate(year = year(CREATED_DATE)) %>%
  select(year, LONGITUDE, LATITUDE) %>%
  filter(year == max(year, na.rm = T)) %>%
  st_as_sf(coords = c("LONGITUDE","LATITUDE")) %>%
  st_set_crs(st_crs(chi_sf))

ggplot() + 
  geom_sf(data = chi_sf) +
  geom_sf(data = c_sf,
    aes(geometry = geometry), 
    color = 'cadetblue',
    alpha = .5,
    stat = "sf_coordinates"
  ) +
  theme_light() + 
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.y=element_blank(),
        axis.title.x=element_blank()) +
  labs(title = '311 Calls', caption = glue::glue('Year: {max(year(max(calls_311$CREATED_DATE)))}')) 


```

### Geographic Coverage: Building Complaints

Highlighting the types of complaints received that mention buildings:

```{r}
calls_311 %>% 
  filter(grepl('Building', SR_TYPE, ignore.case = TRUE)) %>%
  tabyl(SR_TYPE) %>% kable()
  
```

```{r, geo-311-types, warning=FALSE, message=FALSE}

# st_crs(chi_sf)
c_sf <- calls_311 %>% 
  filter(!is.na(LONGITUDE), !is.na(LATITUDE)) %>%
  filter(grepl('Building', SR_TYPE)) %>%
  mutate(year = year(CREATED_DATE)) %>%
  select(year, LONGITUDE, LATITUDE, SR_TYPE) %>%
  filter(year == max(year, na.rm = TRUE)) %>%
  st_as_sf(coords = c("LONGITUDE","LATITUDE")) %>%
  st_set_crs(st_crs(chi_sf))

ggplot() + 
  geom_sf(data = chi_sf) +
  geom_sf(data = c_sf,
    aes(geometry = geometry, 
        color = SR_TYPE),
    stat = "sf_coordinates"
  ) +
  theme_light() + 
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.y=element_blank(),
        axis.title.x=element_blank()) +
  labs(title = '311 Calls: Building Complaints', 
       caption = glue::glue('Year: {max(year(max(calls_311$CREATED_DATE)))}')) 


```

